---
title: "Computational methods for mixed-effects models"
author: "Douglas Bates"
date: "2014-04-16"
output: 
    ioslides_presentation:
        incremental: true
---
```{r preliminaries,cache=FALSE,echo=FALSE,include=FALSE}
library(lattice)
library(knitr)
library(lme4)
opts_chunk$set(cache=TRUE)
options(show.signif.stars=FALSE)
```

# Model definition

## Linear predictors and *effects*

- In statistics, a *linear model* relates a *response* random variable, $\mathcal{Y}$, to observed *covariates* through a *linear predictor*, $\mathbf{X}\mathbf{\beta}$, where $\mathbf{X}$ is a *model matrix*.
$$ \mathcal{Y}\sim\mathcal{N}(\mathbf X\mathbf\beta,\sigma^2\mathbf I) $$

- In a *generalized linear model* (GLM) the mean response is a transformation, called the *link function*, of the linear predictor. For example,
$$\mathcal{Y}\sim\mathcal{Pois}(\exp(\mathbf{X\beta})) .$$

- A categorical covariate (e.g. `sex` or `item`) with $k$ levels generates $k-1$ columns in $\mathbf X$.  The corresponding elements of $\mathbf\beta$ are sometimes called the *effects* of the covariate.

- If the number of levels of the covariate is small and reproducible we use *fixed-effects* coefficients.

## Random effects

- When the levels of the covariate, (e.g. `item`) are a sample from a population of potential values we model the *random effects* as being from a *distribution*.

- The simplest case of a linear model with random effects is
$$ \mathcal{B}\sim\mathcal{N}(\mathbf{0},\sigma^2_b\mathbf{I}),\quad \mathcal{Y}|\mathcal{B}=\mathbf{b}\sim\mathcal{N}(\mathbf{X}\mathbf{\beta}+\mathbf{Z}\mathbf{b},\sigma^2\mathbf I)$$

- Here the *random-effects model matrix*, $\mathbf{Z}$, is the *indicator matrix* of the levels of the covariate.  In general the linear predictor for the conditional distribution $\mathcal{Y}|\mathcal{B}=\mathbf{b}$ becomes $\mathbf{X\beta+Z b}$ (e.g. $\mathcal{Y}|\mathcal{B}=\mathbf{b}\sim\mathcal{Pois}(\exp(\mathbf{X\beta+Z b}))$).

- A **mixed-effects model** has both fixed-effects, $\mathbf{\beta}$, and random effects, $\mathcal{B}$.

- Notice that:
    - it's actually the *levels* that are fixed or random, not the effects
    - in practice any model with random effects is a mixed-effects model.

## Formula language in `lme4` and `MixedModels`

- The [lme4](https://github.com/lme4/lme4) package for [R](http://www.r-project.org) represents the model with a formula.

- The formula language in [Julia](http://julialang.org) was based on that from `R`.  The [MixedModels](https://github.com/dmbates/MixedModels.jl) package for `Julia` uses the same formula representation as does `lme4`.

- Fixed-effects terms in the formula are described in the `lm` [help page](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html).  Random-effects terms are of the form `(f|g)` where
    - `f` is a linear model expression
    - `g` is an expression for a `factor`, called the *grouping factor*
    
- $\mathbf{\Sigma}$, the covariance matrix for the random effects, is block-diagonal with a block for each term.
    - i.e. in the unconditional distribution, $\mathcal{B}\sim\mathcal{N}(\mathbf{0},\mathbf{\Sigma})$, random effects from different terms are independent

- The model matrix $\mathbf{Z}$ is the horizontal concatenation of blocks for each term.

## Simple, scalar random effects

- A *simple, scalar* random-effects term is of the form `(1|g)`.  
- There is one random effect for each level of `g`.  Let $i_g$ be the number of levels.
- Its diagonal block in $\mathbf{\Sigma}$ is $\sigma^2_g\mathbf{I}_{i_g}$.
- Its vertical block in $\mathbf{Z}$ is the $n\times i_g$ indicator matrix for the levels of `g`.

## The `InstEval` data from ETH-Zurich {.build}

- These data are from several years of instructor evaluations at ETH-Zurich. 
    - `y` is the evaluation (larger is better)
    - `s` is the student id
    - `d` is the instructor id.
```{r InstEval}
str(InstEval)
```

## Partially crossed random effects
```{r fm1,cache=TRUE}
system.time(fm1 <- lmer(y ~ dept*service+(1|s)+(1|d),InstEval,REML=FALSE))
VarCorr(fm1)
```

## Fitting LMMs in `Julia`
```{.julia}
using DataFrames,MixedModels
inst = DataFrame(read_rda("./data/InstEval.rda")["InstEval"]);
fm1 = fit(lmm(y ~ dept*service + (1|s) + (1|d), inst))  # next slide
@time fit(lmm(y ~ dept*service + (1|s) + (1|d), inst));
 # elapsed time: 7.170804855 seconds (211949256 bytes allocated)
```

- In `Julia` the `using` directive attaches a package.
- The `read_rda` function reads a saved `R` data set.
- The default criterion for the `lmm` function in the `MixedModels` package is ML.
- The entire `MixedModels` package is written in `Julia`
- Libraries of compiled code such as `OpenBLAS`, `LAPACK` and `SuiteSparse` are used but they are part of the base system.


## Results of model fit
```
Linear mixed model fit by maximum likelihood
 logLik: -118792.776708, deviance: 237585.553415

 Variance components:
                Variance    Std.Dev.
 s              0.105418    0.324681
 d              0.258416    0.508347
 Residual       1.384728    1.176745
 Number of obs: 73421; levels of grouping factors: 2972, 1128

  Fixed-effects parameters:
        Estimate Std.Error   z value
 [1]     3.22961  0.064053   50.4209
 [2]    0.129536  0.101294   1.27882
 [3]   -0.176751 0.0881352  -2.00545
 [4]   0.0517102 0.0817524  0.632522
 [5]   0.0347319  0.085621  0.405647
 [6]     0.14594 0.0997984   1.46235
 [7]    0.151689 0.0816897   1.85689
 [8]    0.104206  0.118751  0.877517
 [9]   0.0440401 0.0962985  0.457329
[10]   0.0517546 0.0986029  0.524879
[11]   0.0466719  0.101942  0.457828
[12]   0.0563461 0.0977925   0.57618
[13]   0.0596536  0.100233   0.59515
[14]  0.00556281  0.110867 0.0501757
[15]    0.252025 0.0686507   3.67112
[16]   -0.180757  0.123179  -1.46744
[17]   0.0186492  0.110017  0.169512
[18]   -0.282269 0.0792937  -3.55979
[19]   -0.494464 0.0790278  -6.25683
[20]   -0.392054  0.110313  -3.55403
[21]   -0.278547 0.0823727  -3.38154
[22]   -0.189526  0.111449  -1.70056
[23]   -0.499868 0.0885423  -5.64553
[24]   -0.497162 0.0917162  -5.42065
[25]    -0.24042 0.0982071   -2.4481
[26]   -0.223013 0.0890548  -2.50422
[27]   -0.516997 0.0809077  -6.38997
[28]   -0.384773  0.091843  -4.18946
```
